---
title: "Stat 651 Project"
author: "Talmage Hilton"
date: "2025-04-08"
output: html_document
---

```{r setup, include=FALSE, waning=FALSE}
library(readxl)
library(tidyverse)
library(dplyr)
library(MASS)
library(coda)
```

```{r}
# Read in data
ratdata <- read_excel("ratdata.xlsx")

# Pivot longer to get tidy data
rat_long <- ratdata %>%
  pivot_longer(cols = starts_with("rat"), names_to = "rat", values_to = "weight")
```



# EDA

```{r, warning=FALSE}
# Weights of each rat
ggplot(rat_long, aes(x = age, y = weight, group = rat, color = rat)) +
  geom_line() +
  geom_point() +
  labs(title = "Growth Curves for Each Rat", x = "Age (days)", y = "Weight (g)") +
  theme_minimal() +
  theme(legend.position = "none")


# Growth rate for each rat
rat_diff <- ratdata %>%
  dplyr::select(-age) %>%
  mutate(across(everything(), ~ c(NA, diff(.)))) %>%
  mutate(age = ratdata$age)

rat_diff_long <- rat_diff %>%
  pivot_longer(cols = -age, names_to = "rat", values_to = "growth_rate")

ggplot(rat_diff_long, aes(x = age, y = growth_rate, group = rat, color = rat)) +
  geom_line() +
  labs(title = "Growth Rate by Age", x = "Age (days)", y = "Δ Weight (g)") +
  theme_minimal() +
  theme(legend.position = "none")


# Boxplots at each age
ggplot(rat_long, aes(x = factor(age), y = weight)) +
  geom_boxplot() +
  labs(title = "Distribution of Rat Weights at Each Age", x = "Age (days)", y = "Weight (g)") +
  theme_minimal()
```

The growth curves show that the rats grow in a mostly linear fashion. Obviously there is some variability, but overall I'd say that it's pretty linear. The growth rates are not very consistent (lots of ups and downs), but overall they're all pretty similar. For the most part the growth rate slows down between days 15-22, then speed up between days 22-29, and then decrease again from days 29-36. However, there are a few rate that increase in growth rate during the last time period. There are some exceptions to the rule here, but overall I'd say that the Normal model for $Y_{ij}$ is fairly reasonable.





# Question 2

```{r}
# Nest the data by rat
rat_nested <- rat_long %>%
  group_by(rat) %>%
  nest()

# Fit regression to each nested dataset
rat_models <- rat_nested %>%
  mutate(model = map(data, ~lm(weight ~ age, data = .)))

# Extract coefficients from each model
rat_coefs <- rat_models %>%
  mutate(coefs = map(model, broom::tidy)) %>%
  unnest(coefs) %>%
  dplyr::select(rat, term, estimate) %>%
  pivot_wider(names_from = term, values_from = estimate) %>%
  rename(alpha_hat = `(Intercept)`, beta_hat = age)


# Check Normality of thetas

# Histogram of intercepts
ggplot(rat_coefs, aes(x = alpha_hat)) +
  geom_histogram(bins = 10, fill = "skyblue") +
  labs(title = "Histogram of Intercepts (alpha_i)", x = expression(hat(alpha)[i]))

# Histogram of slopes
ggplot(rat_coefs, aes(x = beta_hat)) +
  geom_histogram(bins = 10, fill = "lightgreen") +
  labs(title = "Histogram of Slopes (beta_i)", x = expression(hat(beta)[i]))

# QQ plots
qqnorm(rat_coefs$alpha_hat); qqline(rat_coefs$alpha_hat, col = "blue")
qqnorm(rat_coefs$beta_hat); qqline(rat_coefs$beta_hat, col = "blue")

```

I would say that the normality assumption is reasonable here. The distribution of intercepts is approximately normal. The distribution of slopes is a little right skewed, but I wouldn't say it's too egregious. The Q-Q plots also aren't too terrible.





# Gibbs Sampler

```{r}
# Priors
lambda0 <- 0.1
nu0 <- 0.1
eta <- c(0, 0)
Sigma <- matrix(c(10, 0, 0, 10), nrow=2, ncol=2)
C <- matrix(c(5, 0, 0, 5), nrow=2, ncol=2)


set.seed(123)

# Prepare data
Y <- as.matrix(ratdata[, 1:30])
age <- ratdata$age
n <- ncol(Y)  # number of rats = 30
t <- length(age)  # number of time points = 5

# Design matrix for each rat (same across rats)
X_list <- lapply(1:n, function(i) cbind(1, age))
Y_list <- lapply(1:n, function(i) Y[, i])

# Hyperparameters
n_iter <- 10000
burn_in <- 1000

# Storage
theta <- array(NA, dim = c(n, 2, n_iter))  # alpha, beta for each rat
mu <- matrix(NA, nrow = n_iter, ncol = 2)
tau <- numeric(n_iter)

# Initialize
theta_mean <- matrix(0, nrow = n, ncol = 2)
mu[1, ] <- c(0, 0)
tau[1] <- 1

# Gibbs sampler
for (iter in 2:n_iter) {
  # --- 1. Sample theta_i | Y_i, mu, tau ---
  for (i in 1:n) {
    Xi <- X_list[[i]]
    Yi <- Y_list[[i]]
    D_inv <- (1/tau[iter - 1]) * t(Xi) %*% Xi + solve(Sigma)
    D <- solve(D_inv)
    m <- D %*% (tau[iter - 1] * t(Xi) %*% Yi + solve(Sigma) %*% mu[iter - 1, ])
    theta[i, , iter] <- mvrnorm(1, m, D)
  }

  # --- 2. Sample mu | theta, tau ---
  theta_bar <- colMeans(theta[, , iter])
  V_mu_inv <- 30 * solve(Sigma) + solve(C)
  V_mu <- solve(V_mu_inv)
  m_mu <- V_mu %*% (30 * solve(Sigma) %*% theta_bar + solve(C) %*% eta)
  mu[iter, ] <- mvrnorm(1, m_mu, V_mu)

  # --- 3. Sample tau | theta ---
  rss <- 0
  for (i in 1:n) {
    Xi <- X_list[[i]]
    Yi <- Y_list[[i]]
    res <- Yi - Xi %*% theta[i, , iter]
    rss <- rss + sum(res^2)
  }
  shape <- (t * n + nu0) / 2
  rate <- (rss + nu0 * lambda0) / 2
  tau[iter] <- rgamma(1, shape, rate)
}



# POSTERIOR INFERENCE

# Remove burn-in
theta_post <- theta[, , (burn_in+1):n_iter]
mu_post <- mu[(burn_in+1):n_iter, ]
tau_post <- tau[(burn_in+1):n_iter]

# Means and credible intervals
mu_est <- colMeans(mu_post)
mu_ci <- apply(mu_post, 2, quantile, probs = c(0.025, 0.975))

tau_est <- mean(tau_post)
tau_ci <- quantile(tau_post, probs = c(0.025, 0.975))


cat("Posterior Mean of mu_1:\n", mu_est[1], "\n")
cat("Posterior Mean of mu_2:\n", mu_est[2], "\n")
cat("95% CI for mu_1:\n", mu_ci[c(1,2)], "\n")
cat("95% CI for mu_2:\n", mu_ci[c(3,4)], "\n\n")

cat("Posterior Mean of tau:\n", tau_est, "\n")
cat("95% CI for tau:\n", tau_ci, "\n")

# Number of rats and iterations
n_rats <- dim(theta_post)[1]
n_samples <- dim(theta_post)[3]

# Create summary table
theta_summary <- data.frame(
  Rat = 1:n_rats,
  Alpha_Mean = numeric(n_rats),
  Alpha_Lower = numeric(n_rats),
  Alpha_Upper = numeric(n_rats),
  Beta_Mean = numeric(n_rats),
  Beta_Lower = numeric(n_rats),
  Beta_Upper = numeric(n_rats)
)

for (i in 1:n_rats) {
  alpha_samples <- theta_post[i, 1, ]
  beta_samples <- theta_post[i, 2, ]
  
  theta_summary$Alpha_Mean[i] <- mean(alpha_samples)
  theta_summary$Alpha_Lower[i] <- quantile(alpha_samples, 0.025)
  theta_summary$Alpha_Upper[i] <- quantile(alpha_samples, 0.975)
  
  theta_summary$Beta_Mean[i] <- mean(beta_samples)
  theta_summary$Beta_Lower[i] <- quantile(beta_samples, 0.025)
  theta_summary$Beta_Upper[i] <- quantile(beta_samples, 0.975)
}

# View summary
print(theta_summary)
```

```{r}
# CONVERGENCE DIAGNOSTICS


mcmc_mu <- mcmc(mu_post)
mcmc_tau <- mcmc(tau_post)

# Trace plots
traceplot(mcmc_mu[,1], main = "Traceplot of mu_1")
traceplot(mcmc_mu[,2], main = "Traceplot of mu_2")
traceplot(mcmc_tau, main = "Traceplot of tau")

# Autocorrelation
acf(mcmc_tau, main = "ACF of tau")

# Gelman diagnostics (if using multiple chains)
# gelman.diag(as.mcmc.list(list(chain1, chain2)))

# Posterior density
densplot(mcmc_mu[,1], main = "Posterior density of mu[1]")
densplot(mcmc_mu[,2], main = "Posterior density of mu[2]")
densplot(mcmc_tau, main = "Posterior density of tau")




# Get alpha and beta
alpha_post <- theta_post[, 1, ]
beta_post <- theta_post[, 2, ]

# Reshape the alpha and beta samples into a long format for plotting
alpha_df <- data.frame(iteration = rep(1:9000, each = 30),
                       value = as.vector(t(alpha_post)),  # Transpose to stack the columns (observations)
                       observation = rep(1:30, times = 9000),
                       param = "alpha")

beta_df <- data.frame(iteration = rep(1:9000, each = 30),
                      value = as.vector(t(beta_post)),
                      observation = rep(1:30, times = 9000),
                      param = "beta")

# Combine the data frames for plotting
theta_df <- rbind(alpha_df, beta_df)

# Trace plots for alpha
par(mfrow = c(5,2))
for (i in 1:30) {
  # Plot trace for alpha (i-th observation)
  plot(1:9000, alpha_post[i, ], type = "l", 
       main = paste("Trace plot for Alpha (Observation", i, ")"),
       xlab = "Iteration", ylab = "Alpha Value", col = "black")
}

# Trace plots for beta
par(mfrow = c(5,2))
for (i in 1:30) {
  # Plot trace for beta (i-th observation)
  plot(1:9000, beta_post[i, ], type = "l", 
       main = paste("Trace plot for Beta (Observation", i, ")"),
       xlab = "Iteration", ylab = "Beta Value", col = "black")
}




# ACF plots for alpha
par(mfrow = c(5,2))
for (i in 1:30) {
  # ACF for alpha (i-th observation)
  acf(alpha_post[i, ], main = paste("ACF for Alpha (Observation", i, ")"))
}

# ACF plots for beta
par(mfrow = c(5,2))
for (i in 1:30) {
  # ACF for beta (i-th observation)
  acf(beta_post[i, ], main = paste("ACF for Beta (Observation", i, ")"))
}
```


